<<<<<<< Updated upstream
---
title: "Final assignment"
author: "Endre Csikos M2GEGE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Prompt

The final project must be an RMarkdown (or Quarto) document and must contain the following elements:

- Reproducible data reading (preferably from the web)
- Proper textual explanations, and commented code
- Exploratory data analysis with useful visualizations
- Testing of hypothesis by building statistical model(s)
- Verifying that statistical tests are reliable and valid (e.g. residual diagnostics, assumption checks)
- Make sure that all code chunks (including the first) is visible (I.e., set the echo = TRUE)
- The repo must contain a knitted HTML. It is even better if you submit a github page.

Evaluation rubric can be found [here](https://docs.google.com/spreadsheets/d/1sQ5friiHq_BQ8XzS2BXTmTPHLm-LJMezVBADOG8aoW4).

# Data analysis

```{r load in libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidytuesdayR)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(car)
library(broom)
library(pROC)
```

```{r load in the data, message=FALSE, warning=FALSE, include=FALSE}
ttdat <- tidytuesdayR::tt_load(2022, week = 6)
rdat <- as.data.frame(ttdat$airmen)
```


## Data exploration
```{r EDA, message=FALSE, warning=FALSE, include=FALSE}
head(rdat)
str(rdat)
summary(rdat)
unique(rdat$rank_at_graduation)
rdat2<-rdat %>% #recode factors as factors
  mutate(rank_at_graduation = as.factor(rank_at_graduation),
         class = as.factor(class),
         graduated_from = as.factor(graduated_from),
         pilot_type = as.factor(pilot_type),
         military_hometown_of_record = as.factor(military_hometown_of_record),
         state = as.factor(state))
#Itt megvolt az átkódolás, újra ugyanazok a lépések
summary(rdat2)
head(rdat2)
levels(rdat2$class)

rdat2 %>% 
  select()

rdat2 %>% 
  mutate(number_of_aerial_victory_credits=as.factor(round(number_of_aerial_victory_credits)),
         hasvictory = number_of_aerial_victory_credits!=0) %>% 
  ggplot(aes(x=hasvictory,fill=number_of_aerial_victory_credits))+geom_bar(stat="count",position="dodge")+scale_y_log10() # ide kell majd column label az értékekkel

summary(rdat2$rank_at_graduation)

rdat2 %>% #nem volt nagyon lost, csak 2 pilóta
  group_by(reported_lost, reported_lost_date) %>% 
  summarize(n=n())

rdat %>% 
  select(5:10) %>% 
  mutate_all(as.character) %>% 
  rownames_to_column(var="id") %>% 
  pivot_longer(!id, names_to = "names",values_to="values") %>% 
  ggplot(aes(x=values))+facet_wrap(~names, scales="free")+geom_bar()+theme(axis.text.x = element_text(angle=90))

rdat %>% 
  group_by(class, military_hometown_of_record) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))#volt-e olyan osztály, ahol volt valamilyen torzítás a származás szempontjából?

rdat %>% 
  group_by(class) %>% 
  summarize(sum = sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum))

rdat %>% 
  group_by(state) %>% 
  summarize(sum = sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum))

rdat %>% 
  group_by(military_hometown_of_record) %>% 
  summarize(sum=sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum)) %>% 
  head(50) %>% 
  ggplot(aes(x=military_hometown_of_record,y=n))+geom_col()+theme(axis.text.x = element_text(angle=90))

rdat %>% 
  mutate(hasvictory = number_of_aerial_victory_credits!=0) %>% 
  group_by(hasvictory, pilot_type) %>% 
  summarize(n=n()) #Itt jól látszik, hogy csak single engine pilótáknak volt győzelme és közülük is csak egy jelentős kisebbségnek, 10%-nak. A kérdés felmerül: a jelen levő tényezők közül volt-e olyan, ami ehhez hozzájárult?

rdatlong <- rdat %>% 
  select(5:10) %>% 
  mutate_all(as.character) %>% 
  rownames_to_column(var="id") %>% 
  pivot_longer(!id, names_to = "names",values_to="values")

rdatlong$numvic <- NA
for(tempid in unique(rdatlong$id)){
  tempnum <- rdat[tempid,"number_of_aerial_victory_credits"]
  rdatlong[rdatlong$id==tempid,"numvic"] <- tempnum
}

rdatlong %>% 
  group_by(names,values) %>% 
  summarize(summi = sum(numvic)) %>% 
  arrange(desc(summi)) %>% 
  head(20) %>% 
  group_by(names) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))

wizdat <- rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) 
  

rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits), min = min(number_of_aerial_victory_credits), max = max(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0) %>% 
  ggplot(aes(y=meanvic, x=(summi+rnorm(26,0,0.1)), label=class, shape = groupsize, color = gradyear))+geom_point(size=3)+scale_x_continuous(breaks=seq(1,20,1))+xlab("Sum of aerial victories")+ylab("aerial victory number to\nclass size ratio")+theme_minimal()+geom_smooth(method=lm, aes(x=summi, y = meanvic, group = groupsize, linetype = groupsize), inherit.aes=F, se = F)+geom_smooth(method=lm, aes(x=summi, y = meanvic, group = gradyear, color = gradyear), inherit.aes=F, se = F)+scale_y_continuous(breaks=seq(0,1.5,0.1))+labs(caption="Szerintem ezt ne tegyük be, mert nem annyira logikus, hogy van-e értelme.")+theme(plot.caption=element_text(hjust=0))
#+geom_text(alpha=0.4,position="identity", inherit.aes = T, size = 3) > ezt kivettem, mert szerintem nem is kell label erre
#+Viszont szerintem majd ezt tartsuk meg azért, mert azt jól mutatja, hogy pl. az évek során nem lettek sikeresebbek, de valamennyire nagyobbak igen az osztályok
#+Ami még látszik, hogy 
what<-rdat %>% 
  group_by(class) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n()) %>% 
  mutate(meanvic = summi/n) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0) %>% 
  nrow()

wizzdat <- rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits), min = min(number_of_aerial_victory_credits), max = max(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0)

wizzdat %>% 
  ggplot(aes(y=summi, x=(n+rnorm(26,0,0.05)), label=class, shape = groupsize, color = gradyear))+geom_point(size=3)+scale_x_continuous(breaks=seq(0,50,1))+scale_y_continuous(breaks=seq(0,50,1))+geom_smooth(method=lm, inherit.aes = F, aes(x=n, y =summi),se=F,color="black",linetype=2)+geom_smooth(method=lm, inherit.aes = F, aes(x=n, y =summi,color=gradyear),se=F)+xlab("Group size")+ylab("Total number of aerial victories")+ggtitle("Group size ~ Victory sum")+labs(caption="Erős kapcsolat van a végzési év és a csoportméret között\nIgazán kapcsolat a győzelemmel a nagyon kicsi és a nagyobb csoportok esetén van.\nViszont általában véve van egy kapcsolat, csak hogy miért, az a kérdés.")+theme(plot.caption = element_text(hjust = 0))

rdat$classum <- NA
rdat$classize <- NA
rdat$notzero <- NA
for(tempclass in unique(rdat$class)[!is.na(unique(rdat$class))]){
  rdat[rdat$class==tempclass&!is.na(rdat$class),"classum"] <- sum(rdat[rdat$class==tempclass,"number_of_aerial_victory_credits"],na.rm=T)
  rdat[rdat$class==tempclass&!is.na(rdat$class),"classize"] <- nrow(rdat[rdat$class==tempclass&!is.na(rdat$class),])
  rdat[rdat$class==tempclass&!is.na(rdat$class),"notzero"] <- nrow(rdat[rdat$class==tempclass&!is.na(rdat$class)&rdat$number_of_aerial_victory_credits!=0,])
}

rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")),class = as.factor(class), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.05))) %>%
  mutate(class = reorder(class, classum,max, order = is.ordered(rdat$classum)),
         sumfact = cut(classum, breaks = c(0,1,3,5,7,9,11,15), labels = c("1","2-3","4-5","6-7","8-9","10-11","12-15"))) %>% 
  group_by(class, gradyear) %>% 
  filter(classum!=0) %>%
  ggplot(aes(x=class,y=winnum,color=sumfact))+geom_point(size=1,position = position_jitter(width=0.1))+theme(axis.text.x = element_text(angle=90))


rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")),class = as.factor(class), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.05))) %>%
  mutate(class = reorder(class, classize,max, order = is.ordered(rdat$classum)),
         sumfact = cut(classum, breaks = c(0,1,3,5,7,9,11,15), labels = c("1","2-3","4-5","6-7","8-9","10-11","12-15")),
         groupsize = cut(classize, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) %>% 
  group_by(class, gradyear) %>% 
  filter(classum!=0) %>%
  ggplot(aes(x=class,y=winnum,color=groupsize))+geom_point(size=1,position = position_jitter(width=0.1))+
  xlab("Class")+ylab("Personal number of victories")+ggtitle("Personal number of victories in groups ordered by group size")+labs(caption="Itt látszik, hogy mintha a csoportméret növekedésével kevesebb lenne az igazán sikeres pilóta.\nLehet, hogy ott már kevésbé figyeltek oda rájuk?")+theme(plot.caption=element_text(hjust=0))+theme_minimal()+theme(axis.text.x = element_text(angle=90))

rdat %>% #Class size ~ number of participants with victory; color = victory number groups
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")),class = as.factor(class), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.05))) %>%
  mutate(class = reorder(class, classize,max, order = is.ordered(rdat$classum)),
         sumfact = cut(classum, breaks = c(0,1,3,5,7,9,11,15), labels = c("1","2-3","4-5","6-7","8-9","10-11","12-15")),
         groupsize = cut(classize, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60")),
         groupwin = cut(classum, breaks = c(0,2,4,7,15), labels = c("0-2","3-4","5-7","8-15"))) %>% 
  group_by(class, gradyear,groupsize,groupwin,classize) %>%
  summarize(max=max(classum),notzero=sum(number_of_aerial_victory_credits>0)) %>% 
  arrange(desc(max)) %>%
  filter(max>0) %>% 
  ggplot(aes(x=classize,y=notzero,color=groupwin))+geom_point()+scale_x_continuous(breaks=seq(1,40,1))+geom_smooth(method=lm,inherit.aes=F,aes(x=classize,y=notzero),se=F)+
  theme_classic()+ggtitle("Number of participants with victory along the size of the class")+xlab("Class size")+ylab("Number of participants with victory")+
  scale_y_continuous(breaks=c(1:8))+labs(color = "Number of victories per class")+labs(caption = "Number of participants with at least on victory seems to increase\nwith the size of the class.")+
  theme(plot.caption = element_text(hjust=0, size = 16))

grady<-rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")),class = as.factor(class), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.05))) %>%
  mutate(class = reorder(class, classize,max, order = is.ordered(rdat$classum)),
         sumfact = cut(classum, breaks = c(0,1,3,5,7,9,11,15), labels = c("1","2-3","4-5","6-7","8-9","10-11","12-15")),
         groupsize = cut(classize, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60")),
         groupwin = cut(classum, breaks = c(0,2,4,7,15), labels = c("0-2","3-4","5-7","8-15"))) %>% 
  group_by(class, gradyear,groupsize,groupwin,classize) %>%
  summarize(max=max(classum),notzero=sum(number_of_aerial_victory_credits>0)) %>% 
  mutate(ratio = (max/notzero)*(notzero/classize)) %>% 
  arrange(desc(ratio)) %>%
  filter(max>0)

grady %>% #ezzel lehet megmutatni azt, hogy a gradyear az, ami alapvetően számít a teljes classra nézve és nem a class létszáma
  ggplot(aes(x=classize,y=ratio,color=gradyear))+geom_point()+geom_hline(yintercept = 0.3)+theme_classic()+ggtitle("Victory ratio along class size")+xlab("Class size (Number of pilots)")+ylab("Victory ratio")+labs(caption="Victory ratio was calculated the following way: (Sum of group victory / Number of pilots with victory)\n multiplied by (Number of pilots with victory / Class size)",color="Graduation year")+theme(plot.caption = element_text(hjust=0))
#szerintem itt úgy kellene módosítani, hogy a labeleket a victory number vagy a csoportnagyság alapján sorba rendezni?

#Konklúzió: nincs strength in numbers, mert ahogy nő átlagosan a csoportméret, úgy csökken az arány. Van egy gyengébb kapcsolat az évekkel, viszont ez valószínűleg inkább ahhoz kapcsolódik, hogy kisebbek voltak a csoportok átlagosan. Ezt egy summaryvel is jó lenne megmutatni.
#+Amit szerintem még érdemes lenne megnézni:
#+Csoportméret~győzelemszámsum [X]
#+évszám~győzelemszámsum []
#+facet_wrap: classokra lebontva, egyéni num of victories []
#
#Amiket pedig summaryvel lenne fontos megmutatni:
#+minden változónál a mintázatokat
#+faktoroknál azt, hogy mondjuk éves szinten hogyan oszlottak meg

#Illetve még egy hiányzó plotot majd később, ha megvan a longformat: azt, hogy a győzelmek hogyan oszlottak még évszámok szerint.

```

### Codebook --> ezt majd valahol az elején kellene bemutatni szerintem.

- name - full name [X]
- last_name **no need to modify** [X]
- first_name **no need to modify** [X]
- graduation_date - this is in POSIXct, good to use **create new var: graduation year** [X]
- rank_at_graduation - factor, recode "Captain" to capt. NA and Unk could be thought of as different levels, keep them. Convert "N/A" to actual NA values.
- class - factor, the id of the class. It represents different levels of pilot training. pilot_type-graduation_year-graduation_month. Technically it is an aggregate of multiple different informations.
- pilot_type - factor, sing
- military_hometown_of_record - Factor, military hometown where the pilots were based
- state - factor
- aerial_victory_credits - char, contains info about the downed plane and the date of the victory. *split up for later use!*
- number_of_aerial_victory_credits - int, contains the number of the downed enemy plane
- reported_lost
- reported_lost_date
- reported_lost_location
- web_profile

```{r EDA datwiz}

```


*Note*
- aerial victories have to be split up according to semicolons.
- pilot_type has to be corrected.



**Optional** splitting could also include the downed plane, could look for additional info about the specific planes.
**Optional** graduation date -> meg lehetne nézni, hogy akik ugyanakkor végeztek, azok hasonlóan álltak győzelmi pontok terén. Ezt igazából talán tovább is lehetne bontani és megnézni, hogy akik valamilyen hónapban végeztek, azok esetleg jobban teljesítettek.
**Optional** Az pl. egy jó hipotézis, hogy aki twin engine képzést kapott, az kevesebb győzelemmel rendelkezik, mert az a B-25 Mitchell közepes bombázót jelentette.

## Data modification

```{r Correct variables}
rdat2<-rdat %>% #recode factors as factors
  mutate(rank_at_graduation = as.factor(rank_at_graduation),
         class = as.factor(class),
         graduated_from = as.factor(graduated_from),
         pilot_type = as.factor(pilot_type),
         military_hometown_of_record = as.factor(military_hometown_of_record),
         state = as.factor(state))
rdat2 <- rdat2 %>% 
  mutate(rank_at_graduation = recode(rank_at_graduation, "Captain" = "Capt", "Unk" = "N/A"), pilot_type = recode(pilot_type, "Liason pilot" = "Liaison pilot"))
#create binary has victory
rdat2 <- rdat2 %>% 
  mutate(hasvictory = number_of_aerial_victory_credits!=0, gradyear = as.numeric(format(graduation_date, format = "%Y"))) 
rdat2[rdat2$hasvictory==T,"hasvictory"] <- 1
rdat2[rdat2$hasvictory==F,"hasvictory"] <- 0
#rdat2<-rdat2 %>% 
#  filter(gradyear<=1944) #mindenkit kidobni, aki később végzett, mert túl későn érkeztek a háborúba
#correct victories manually
rdat2[495,"aerial_victory_credits"] <- "Downed 1 Me-109 on June 9, 1944; Downed 1 He-111 on October 12, 1944; Downed 1 Me-109 on October 12, 1944"
rdat2[240,"aerial_victory_credits"] <- "Downed 1 Fw-190 on July 2, 1943; Downed 1 Fw-190 on January 28, 1944; Downed 1 Me-109 on January 28, 1944"
#Throw those without any info about graduation year
rdat2<-rdat2[!is.na(rdat2$class)|!is.na(rdat2$graduation_date),]
#Drop without rank
rdat2 <- rdat2[!is.na(rdat2$rank_at_graduation),]
#Correct missing gradyear
rdat2[is.na(rdat2$gradyear),"gradyear"] <- 1945
```


```{r Create new variables}

#####Create long format with the victories --> ezt majd szerintem szét kellene szedni külön 2 chunkba
row.names(rdat2) <- 1:nrow(rdat2)
rlong <- as.data.frame(matrix(nrow=0,ncol=18))
#colnames(rlong) <- c("id","num","type","date")
cn <- c(c("id","num","type","date"),colnames(rdat2[,c(4:10,12:18)]))
colnames(rlong) <- cn
for(temprow in (1:nrow(rdat2))){
  temps <- rdat2[temprow,c(4:10,12:18)]
  tempid <- temprow
  tempvic <- rdat2[temprow,"aerial_victory_credits"]
  tempnum <- NA
  temptype <- NA
  tempdate <- NA
  if(!is.na(tempvic)){
    tempvic2 <- strsplit(tempvic, split="; ")
    for(vic in unlist(tempvic2)){
      vic2<-unlist(strsplit(unlist(vic),split=" "))
      if(length(vic2)==8){
        tempnum1 <- eval(parse(text=vic2[2]))
        tempnum2 <- eval(parse(text=vic2[3]))
        tempnum <- tempnum1 + tempnum2
        vic2 <- vic2[-3]
      }else{
        tempnum <- eval(parse(text=vic2[2]))
      }
      temptype <- vic2[3]
      tempdate <- as.Date(gsub(",","",paste(vic2[7], vic2[5], vic2[6],sep="-")), format = "%Y-%B-%d")
      temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
      colnames(temprow) <- cn
      rlong <- rbind(rlong,temprow)
    }
  }else{
    temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
    colnames(temprow) <- cn
    rlong <- rbind(rlong,temprow)
  }
}
rlong$date <- as.Date(rlong$date)
```

```{r long format data exploration}
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  group_by(date) %>% 
  summarize(n=n())
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1),y=as.numeric(format(date, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+scale_y_continuous(breaks = seq(1941,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
    ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.1),y=number_of_aerial_victory_credits+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.05)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
  group_by(gradyear,rank_at_graduation,state) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))

rlong %>% #Number of victories per motnh
  mutate(shotyear = format(date,format="%Y-%m")) %>% 
  group_by(shotyear) %>% 
  summarize(number=n()) %>% 
  mutate(shotyear = as.factor(shotyear)) %>% 
  filter(!is.na(shotyear)) %>% 
  ggplot(aes(x=shotyear,y=number))+geom_point()+theme_minimal()+theme(axis.text.x = element_text(angle=-45))+xlab("Date of victory")+ylab("Number of victories")+ggtitle("Number of victories per months")+
  scale_y_continuous(breaks=seq(0,35,2))
```

## Model building

### Logistic regression

```{r}
logmod <-  glm(hasvictory ~ rank_at_graduation + gradyear, data = rdat2, family = binomial(link = "logit")) #itt szerintem azért kérdés, hogy mennyire van értelme az 1945-ös csoportot benntartani, szerintem semmi. Nem is nagyon lehet megérveli. A pilot_type pedig nyilván nem kell, mert ott csak a single engine releváns.
binommod <- MASS::glm.nb(number_of_aerial_victory_credits~rank_at_graduation+gradyear, data = rdat2)
str(rdat2)
results <- list(summary(logmod),summary(binommod))
```

#### Assumption checks

##### Logistic regression

Assumption tests for the logistic regression model, including linear connection between outcome and continuous predictor variables, multicollinearity and influential cases. Testing for linear association between the continuous predictor of graduation year and the outcome variable.

```{r linear connection}
rdat2test <- rdat2
rdat2test$gradyear_log <- rdat2test$gradyear * log(rdat2test$gradyear)
tester<-glm(hasvictory ~ gradyear + gradyear_log,data=rdat2test,family=binomial)
summary(tester) #linear connection is violated, transforming gradyear to categorical variable.
```

The linear connection between the logit and graduation year is violated, thus graduation year needs to be converted to a categorical variable to circumvent the issue.

```{r recode gradyear to categorical}
rdat2$gradyear <- factor(rdat2$gradyear, levels = c(1942:1948),ordered=T)
logmod <-  glm(hasvictory ~ rank_at_graduation + gradyear, data = rdat2, family = binomial(link = "logit")) #rebuild the model after recoding gradyear
```

Testing for multicollinearity.

```{r multicollinearity}

vif(logmod)
```

VIF suggests the absence of multicollinearity

```{r influential cases}
logmod_stat <- augment(logmod)
plot(logmod,which=4,id.n=3)
```

Cook's distance strict threshold of 4/N (`r 4/nrow(rdat2)`) is violated here, but since the most influential cases are under 1, it is not likely that they have a strong influence on our model.

###### Summary

With the absence of multicollinearity and influential cases, and with our outcome variable being binary, it seems that the use of a logistic regression is supported.

##### Negative binomial regression

Assumption tests for the logistic regression model. Compared to the logistic regression we are going to additionally test for the overdispersion to make sure that a Poisson count model would not be more sufficient for our data.

```{r}
mean(rdat2$number_of_aerial_victory_credits, na.rm=T)
var(rdat2$number_of_aerial_victory_credits, na.rm=T)
poisson <- glm(number_of_aerial_victory_credits~gradyear+rank_at_graduation, data = rdat2, family=poisson)
nb <- MASS::glm.nb(number_of_aerial_victory_credits~gradyear+rank_at_graduation, data = rdat2)
lmtest::lrtest(poisson,nb)
```

Likelihood Ratio Test suggest overdispersion, negative binomial model is superior to a Poisson count model.

# Report

## Model results

### Binary logistic

```{r}
summary(logmod)
Anova(logmod, type = "II", test = "LR")
```

Overall in the model the individual years as predictors don't have a significant effect, graduation year itself seems to be a significant predictor based on the Analysis of Deviance. #ide majd kellene valami hivatkozás, hogy miért ezt futtatom.

```{r}
summary(nb)
Anova(nb, type = "II", test = "LR")
```


## Model comparison

To compare our two models on how well they predict a pilot's victory.

```{r}
logmod_pred <- predict(logmod, type = "response")
mu_nb <- fitted(nb)
theta_nb <- nb$theta
probzero_nb <- dnbinom(0, mu=mu_nb, size=theta_nb)
probnotzero_nb <- 1 - probzero_nb
roc_log <- roc(rdat2$hasvictory, logmod_pred)
roc_nb <- roc(rdat2$hasvictory, probnotzero_nb)
cat("logistic model AUC:",auc(roc_log),"\n",sep=" ")
cat("negative binom model AUC:",auc(roc_nb),"\n",sep=" ")
```

Both models yielded an identical AUC of 0.76, which suggests that the choice of distribution (binary vs identical) did not alter the rank-order of the pilots.

=======
---
title: "Final assignment"
author: "Endre Csikos M2GEGE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Prompt

The final project must be an RMarkdown (or Quarto) document and must contain the following elements:

- Reproducible data reading (preferably from the web)
- Proper textual explanations, and commented code
- Exploratory data analysis with useful visualizations
- Testing of hypothesis by building statistical model(s)
- Verifying that statistical tests are reliable and valid (e.g. residual diagnostics, assumption checks)
- Make sure that all code chunks (including the first) is visible (I.e., set the echo = TRUE)
- The repo must contain a knitted HTML. It is even better if you submit a github page.

Evaluation rubric can be found [here](https://docs.google.com/spreadsheets/d/1sQ5friiHq_BQ8XzS2BXTmTPHLm-LJMezVBADOG8aoW4).

# Data analysis

```{r load in libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidytuesdayR)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r load in the data, message=FALSE, warning=FALSE, include=FALSE}
ttdat <- tidytuesdayR::tt_load(2022, week = 6)
rdat <- as.data.frame(ttdat$airmen)
```


## Data exploration
```{r EDA, message=FALSE, warning=FALSE, include=FALSE}
head(rdat)
str(rdat)
summary(rdat)
unique(rdat$rank_at_graduation)
rdat2<-rdat %>% #recode factors as factors
  mutate(rank_at_graduation = as.factor(rank_at_graduation),
         class = as.factor(class),
         graduated_from = as.factor(graduated_from),
         pilot_type = as.factor(pilot_type),
         military_hometown_of_record = as.factor(military_hometown_of_record),
         state = as.factor(state))
#Itt megvolt az átkódolás, újra ugyanazok a lépések
summary(rdat2)
head(rdat2)
levels(rdat2$class)

rdat2 %>% 
  select()

rdat2 %>% 
  mutate(number_of_aerial_victory_credits=as.factor(round(number_of_aerial_victory_credits)),
         hasvictory = number_of_aerial_victory_credits!=0) %>% 
  ggplot(aes(x=hasvictory,fill=number_of_aerial_victory_credits))+geom_bar(stat="count",position="dodge")+scale_y_log10() # ide kell majd column label az értékekkel

summary(rdat2$rank_at_graduation)

rdat2 %>% #nem volt nagyon lost, csak 2 pilóta
  group_by(reported_lost, reported_lost_date) %>% 
  summarize(n=n())

rdat %>% 
  select(5:10) %>% 
  mutate_all(as.character) %>% 
  rownames_to_column(var="id") %>% 
  pivot_longer(!id, names_to = "names",values_to="values") %>% 
  ggplot(aes(x=values))+facet_wrap(~names, scales="free")+geom_bar()+theme(axis.text.x = element_text(angle=90))

rdat %>% 
  group_by(class, military_hometown_of_record) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))#volt-e olyan osztály, ahol volt valamilyen torzítás a származás szempontjából?

rdat %>% 
  group_by(class) %>% 
  summarize(sum = sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum))

rdat %>% 
  group_by(state) %>% 
  summarize(sum = sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum))

rdat %>% 
  group_by(military_hometown_of_record) %>% 
  summarize(sum=sum(number_of_aerial_victory_credits),n=n()) %>% 
  arrange(desc(sum)) %>% 
  head(50) %>% 
  ggplot(aes(x=military_hometown_of_record,y=n))+geom_col()+theme(axis.text.x = element_text(angle=90))

rdat %>% 
  mutate(hasvictory = number_of_aerial_victory_credits!=0) %>% 
  group_by(hasvictory, pilot_type) %>% 
  summarize(n=n()) #Itt jól látszik, hogy csak single engine pilótáknak volt győzelme és közülük is csak egy jelentős kisebbségnek, 10%-nak. A kérdés felmerül: a jelen levő tényezők közül volt-e olyan, ami ehhez hozzájárult?

rdatlong <- rdat %>% 
  select(5:10) %>% 
  mutate_all(as.character) %>% 
  rownames_to_column(var="id") %>% 
  pivot_longer(!id, names_to = "names",values_to="values")

rdatlong$numvic <- NA
for(tempid in unique(rdatlong$id)){
  tempnum <- rdat[tempid,"number_of_aerial_victory_credits"]
  rdatlong[rdatlong$id==tempid,"numvic"] <- tempnum
}

rdatlong %>% 
  group_by(names,values) %>% 
  summarize(summi = sum(numvic)) %>% 
  arrange(desc(summi)) %>% 
  head(20) %>% 
  group_by(names) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))

wizdat <- rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) 
  

rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits), min = min(number_of_aerial_victory_credits), max = max(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0) %>% 
  ggplot(aes(y=meanvic, x=(summi+rnorm(26,0,0.1)), label=class, shape = groupsize, color = gradyear))+geom_point(size=3)+scale_x_continuous(breaks=seq(1,20,1))+xlab("Sum of aerial victories")+ylab("aerial victory number to\nclass size ratio")+theme_minimal()+geom_smooth(method=lm, aes(x=summi, y = meanvic, group = groupsize, linetype = groupsize), inherit.aes=F, se = F)+geom_smooth(method=lm, aes(x=summi, y = meanvic, group = gradyear, color = gradyear), inherit.aes=F, se = F)+scale_y_continuous(breaks=seq(0,1.5,0.1))
#+geom_text(alpha=0.4,position="identity", inherit.aes = T, size = 3) > ezt kivettem, mert szerintem nem is kell label erre
#+Viszont szerintem majd ezt tartsuk meg azért, mert azt jól mutatja, hogy pl. az évek során nem lettek sikeresebbek, de valamennyire nagyobbak igen az osztályok
#+Ami még látszik, hogy 
what<-rdat %>% 
  group_by(class) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n()) %>% 
  mutate(meanvic = summi/n) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0) %>% 
  nrow()

wizzdat <- rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y"))) %>% 
  group_by(class, gradyear) %>% 
  summarize(summi = sum(number_of_aerial_victory_credits),n=n(), sd = sd(number_of_aerial_victory_credits), min = min(number_of_aerial_victory_credits), max = max(number_of_aerial_victory_credits)) %>% 
  mutate(meanvic = summi/n, groupsize = cut(n, breaks = c(0,10,20,30,40,50,60), labels = c("0-10","11-20","21-30","31-40","41-50","51-60"))) %>% 
  arrange(desc(meanvic)) %>% 
  filter(summi!=0)

wizzdat %>% 
  ggplot(aes(y=summi, x=(n+rnorm(26,0,0.05)), label=class, shape = groupsize, color = gradyear))+geom_point(size=3)+scale_x_continuous(breaks=seq(0,50,1))+scale_y_continuous(breaks=seq(0,50,1))+geom_smooth(method=lm, inherit.aes = F, aes(x=n, y =summi),se=F,color="black",linetype=2)+geom_smooth(method=lm, inherit.aes = F, aes(x=n, y =summi,color=gradyear),se=F)

rdat$sumclass <- NA
for(klasz in unique(rdat[!is.na(rdat$class),"class"])){
 rdat[rdat$class==klasz&!is.na(rdat$class),"sumclass"] <- sum(rdat[rdat$class==klasz,"number_of_aerial_victory_credits"],na.rm=T) 
}
  

rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.1)),
         ,class = as.factor(class)) %>% 
  group_by(class, gradyear) %>% 
  filter(sumclass!=0) %>% 
  ggplot(aes(x=as.numeric(class)+rnorm(length(class),0,0.1),y=winnum))+geom_point()+theme(axis.text.x = element_text(angle=90))+scale_x_continuous(breaks=seq(1,length(labels(as.factor(rdat$class))),1), labels = labels(class)) #szerintem itt úgy kellene módosítani, hogy a labeleket a csoportszám alapján kellene sorba rendezni.

rdat %>% 
  mutate(gradyear = as.factor(format(graduation_date, format = "%Y")), winnum=(number_of_aerial_victory_credits+rnorm(length(number_of_aerial_victory_credits),0,0.1)),
         ,class = as.factor(class)) %>% 
  group_by(class, gradyear) %>% 
  filter(sumclass!=0) %>% 
  length(unique(class))

#Konklúzió: nincs strength in numbers, mert ahogy nő átlagosan a csoportméret, úgy csökken az arány. Van egy gyengébb kapcsolat az évekkel, viszont ez valószínűleg inkább ahhoz kapcsolódik, hogy kisebbek voltak a csoportok átlagosan. Ezt egy summaryvel is jó lenne megmutatni.
#+Amit szerintem még érdemes lenne megnézni:
#+Csoportméret~győzelemszámsum
#+évszám~győzelemszámsum
#+facet_wrap: classokra lebontva, egyéni num of victories
#
#Amiket pedig summaryvel lenne fontos megmutatni:
#+minden változónál a mintázatokat
#+faktoroknál azt, hogy mondjuk éves szinten hogyan oszlottak meg

#Illetve még egy hiányzó plotot majd később, ha megvan a longformat: azt, hogy a győzelmek hogyan oszlottak még évszámok szerint.

```

### Codebook --> ezt majd valahol az elején kellene bemutatni szerintem.

- name - full name [X]
- last_name **no need to modify** [X]
- first_name **no need to modify** [X]
- graduation_date - this is in POSIXct, good to use **create new var: graduation year** [X]
- rank_at_graduation - factor, recode "Captain" to capt. NA and Unk could be thought of as different levels, keep them. Convert "N/A" to actual NA values.
- class - factor, the id of the class. It represents different levels of pilot training. pilot_type-graduation_year-graduation_month. Technically it is an aggregate of multiple different informations.
- pilot_type - factor, sing
- military_hometown_of_record - Factor, military hometown where the pilots were based
- state - factor
- aerial_victory_credits - char, contains info about the downed plane and the date of the victory. *split up for later use!*
- number_of_aerial_victory_credits - int, contains the number of the downed enemy plane
- reported_lost
- reported_lost_date
- reported_lost_location
- web_profile

```{r EDA datwiz}

```


*Note*
- aerial victories have to be split up according to semicolons.
- pilot_type has to be corrected.



**Optional** splitting could also include the downed plane, could look for additional info about the specific planes.
**Optional** graduation date -> meg lehetne nézni, hogy akik ugyanakkor végeztek, azok hasonlóan álltak győzelmi pontok terén. Ezt igazából talán tovább is lehetne bontani és megnézni, hogy akik valamilyen hónapban végeztek, azok esetleg jobban teljesítettek.
**Optional** Az pl. egy jó hipotézis, hogy aki twin engine képzést kapott, az kevesebb győzelemmel rendelkezik, mert az a B-25 Mitchell közepes bombázót jelentette.

## Data modification

```{r Correct variables}
rdat2 <- rdat2 %>% 
  mutate(rank_at_graduation = recode(rank_at_graduation, "Captain" = "Capt", "Unk" = "N/A"), pilot_type = recode(pilot_type, "Liason pilot" = "Liaison pilot"))
rdat2 <- rdat2 %>% 
  mutate(hasvictory = number_of_aerial_victory_credits!=0, gradyear = as.numeric(format(graduation_date, format = "%Y"))) 
rdat2<-rdat2 %>% 
  filter(gradyear<=1944) #mindenkit kidobni, aki később végzett, mert túl későn érkeztek a háborúba
rdat2[495,"aerial_victory_credits"] <- "Downed 1 Me-109 on June 9, 1944; Downed 1 He-111 on October 12, 1944; Downed 1 Me-109 on October 12, 1944"
rdat2[240,"aerial_victory_credits"] <- "Downed 1 Fw-190 on July 2, 1943; Downed 1 Fw-190 on January 28, 1944; Downed 1 Me-109 on January 28, 1944"
```


```{r Create new variables}

#####Create long format with the victories --> ezt majd szerintem szét kellene szedni külön 2 chunkba
row.names(rdat2) <- 1:nrow(rdat2)
rlong <- as.data.frame(matrix(nrow=0,ncol=18))
#colnames(rlong) <- c("id","num","type","date")
cn <- c(c("id","num","type","date"),colnames(rdat2[,c(4:10,12:18)]))
colnames(rlong) <- cn
for(temprow in (1:nrow(rdat2))){
  temps <- rdat2[temprow,c(4:10,12:18)]
  tempid <- temprow
  tempvic <- rdat2[temprow,"aerial_victory_credits"]
  tempnum <- NA
  temptype <- NA
  tempdate <- NA
  if(!is.na(tempvic)){
    tempvic2 <- strsplit(tempvic, split="; ")
    for(vic in unlist(tempvic2)){
      vic2<-unlist(strsplit(unlist(vic),split=" "))
      if(length(vic2)==8){
        tempnum1 <- eval(parse(text=vic2[2]))
        tempnum2 <- eval(parse(text=vic2[3]))
        tempnum <- tempnum1 + tempnum2
        vic2 <- vic2[-3]
      }else{
        tempnum <- eval(parse(text=vic2[2]))
      }
      temptype <- vic2[3]
      tempdate <- as.Date(gsub(",","",paste(vic2[7], vic2[5], vic2[6],sep="-")), format = "%Y-%B-%d")
      temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
      colnames(temprow) <- cn
      rlong <- rbind(rlong,temprow)
    }
  }else{
    temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
    colnames(temprow) <- cn
    rlong <- rbind(rlong,temprow)
  }
}
rlong$date <- as.Date(rlong$date)
```

```{r long format data exploration}
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  group_by(date) %>% 
  summarize(n=n())
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1),y=as.numeric(format(date, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+scale_y_continuous(breaks = seq(1941,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
    ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.1),y=number_of_aerial_victory_credits+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.05)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
  group_by(gradyear,rank_at_graduation,state) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

# Report

>>>>>>> Stashed changes
# Discussion