---
title: "Final assignment"
author: "Endre Csikos M2GEGE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Prompt

The final project must be an RMarkdown (or Quarto) document and must contain the following elements:

- Reproducible data reading (preferably from the web)
- Proper textual explanations, and commented code
- Exploratory data analysis with useful visualizations
- Testing of hypothesis by building statistical model(s)
- Verifying that statistical tests are reliable and valid (e.g. residual diagnostics, assumption checks)
- Make sure that all code chunks (including the first) is visible (I.e., set the echo = TRUE)
- The repo must contain a knitted HTML. It is even better if you submit a github page.

Evaluation rubric can be found [here](https://docs.google.com/spreadsheets/d/1sQ5friiHq_BQ8XzS2BXTmTPHLm-LJMezVBADOG8aoW4).

# Data analysis

```{r load in libraries and data, message=FALSE, warning=FALSE, include=FALSE}
library(tidytuesdayR)
library(dplyr)
library(tidyverse)
ttdat <- tidytuesdayR::tt_load(2022, week = 6)
rdat <- as.data.frame(ttdat$airmen)
```

## Data exploration
```{r EDA, message=FALSE, warning=FALSE, include=FALSE}
str(rdat)
unique(rdat$rank_at_graduation)
rdat2<-rdat %>% 
  mutate(rank_at_graduation = as.factor(rank_at_graduation),
         class = as.factor(class),
         graduated_from = as.factor(graduated_from),
         pilot_type = as.factor(pilot_type),
         military_hometown_of_record = as.factor(military_hometown_of_record),
         state = as.factor(state))
str(rdat2)
summary(rdat2)
head(rdat2)
```

### Codebook

- name - full name
- last_name
- first_name
- graduation_date - this is in POSIXct, good to use
- rank_at_graduation - factor
- class - factor, the id of the class. It represents different levels of pilot training.
- pilot_type - factor, sing
- military_hometown_of_record - Factor, military hometown where the pilots were based
- state - factor
- aerial_victory_credits - char, contains info about the downed plane and the date of the victory. *split up for later use!*
- number_of_aerial_victory_credits - int, contains the number of the downed enemy plane
- reported_lost
- reported_lost_date
- reported_lost_location
- web_profile


*Note*
- aerial victories have to be split up according to semicolons.
- pilot_type has to be corrected.



**Optional** splitting could also include the downed plane, could look for additional info about the specific planes.



# Report

# Discussion