---
title: "Final assignment"
author: "Endre Csikos M2GEGE"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Prompt

The final project must be an RMarkdown (or Quarto) document and must contain the following elements:

- Reproducible data reading (preferably from the web)
- Proper textual explanations, and commented code
- Exploratory data analysis with useful visualizations
- Testing of hypothesis by building statistical model(s)
- Verifying that statistical tests are reliable and valid (e.g. residual diagnostics, assumption checks)
- Make sure that all code chunks (including the first) is visible (I.e., set the echo = TRUE)
- The repo must contain a knitted HTML. It is even better if you submit a github page.

Evaluation rubric can be found [here](https://docs.google.com/spreadsheets/d/1sQ5friiHq_BQ8XzS2BXTmTPHLm-LJMezVBADOG8aoW4).

# Data analysis

```{r load in libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidytuesdayR)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r load in the data, message=FALSE, warning=FALSE, include=FALSE}
ttdat <- tidytuesdayR::tt_load(2022, week = 6)
rdat <- as.data.frame(ttdat$airmen)
```


## Data exploration
```{r EDA, message=FALSE, warning=FALSE, include=FALSE}
str(rdat)
unique(rdat$rank_at_graduation)
rdat2<-rdat %>% 
  mutate(rank_at_graduation = as.factor(rank_at_graduation),
         class = as.factor(class),
         graduated_from = as.factor(graduated_from),
         pilot_type = as.factor(pilot_type),
         military_hometown_of_record = as.factor(military_hometown_of_record),
         state = as.factor(state))
str(rdat2)
summary(rdat2)
head(rdat2)
levels(rdat2$class)
rdat2 %>% 
  filter(class=="45-BI-1") %>% 
  select(pilot_type) %>% 
  summary() #mit jelent ez a class? A type szerint service pilot. AI szerint nem volt ilyen designation.


rdat2 %>% 
  mutate(number_of_aerial_victory_credits = (number_of_aerial_victory_credits + rnorm(nrow(rdat2), 0, 0.1))) %>% 
  ggplot(aes(y=number_of_aerial_victory_credits,x=rnorm(nrow(rdat2),1,0.2)))+geom_point()+geom_hline(yintercept=0.5) #itt látszik jól, hogy a többségnek egyáltalán nem volt légi győzelme.

rdat2 %>% #nem volt nagyon lost
  group_by(reported_lost, reported_lost_date) %>% 
  summarize(n=n())

rdat2 %>% #2nd Lt volt a legtöbb, akinek volt légi győzelme
  group_by(pilot_type, hasvictory, rank_at_graduation) %>% 
  summarize(n=n())


```

### Codebook --> ezt majd valahol az elején kellene bemutatni szerintem.

- name - full name [X]
- last_name **no need to modify** [X]
- first_name **no need to modify** [X]
- graduation_date - this is in POSIXct, good to use **create new var: graduation year** [X]
- rank_at_graduation - factor, recode "Captain" to capt. NA and Unk could be thought of as different levels, keep them. Convert "N/A" to actual NA values.
- class - factor, the id of the class. It represents different levels of pilot training. pilot_type-graduation_year-graduation_month. Technically it is an aggregate of multiple different informations.
- pilot_type - factor, sing
- military_hometown_of_record - Factor, military hometown where the pilots were based
- state - factor
- aerial_victory_credits - char, contains info about the downed plane and the date of the victory. *split up for later use!*
- number_of_aerial_victory_credits - int, contains the number of the downed enemy plane
- reported_lost
- reported_lost_date
- reported_lost_location
- web_profile


*Note*
- aerial victories have to be split up according to semicolons.
- pilot_type has to be corrected.



**Optional** splitting could also include the downed plane, could look for additional info about the specific planes.
**Optional** graduation date -> meg lehetne nézni, hogy akik ugyanakkor végeztek, azok hasonlóan álltak győzelmi pontok terén. Ezt igazából talán tovább is lehetne bontani és megnézni, hogy akik valamilyen hónapban végeztek, azok esetleg jobban teljesítettek.
**Optional** Az pl. egy jó hipotézis, hogy aki twin engine képzést kapott, az kevesebb győzelemmel rendelkezik, mert az a B-25 Mitchell közepes bombázót jelentette.

## Data modification

```{r Correct variables}
rdat2 <- rdat2 %>% 
  mutate(rank_at_graduation = recode(rank_at_graduation, "Captain" = "Capt", "Unk" = "N/A"), pilot_type = recode(pilot_type, "Liason pilot" = "Liaison pilot"))
rdat2 <- rdat2 %>% 
  mutate(hasvictory = number_of_aerial_victory_credits!=0, gradyear = as.numeric(format(graduation_date, format = "%Y"))) 
rdat2<-rdat2 %>% 
  filter(gradyear<=1944) #mindenkit kidobni, aki később végzett, mert túl későn érkeztek a háborúba
rdat2[495,"aerial_victory_credits"] <- "Downed 1 Me-109 on June 9, 1944; Downed 1 He-111 on October 12, 1944; Downed 1 Me-109 on October 12, 1944"
rdat2[240,"aerial_victory_credits"] <- "Downed 1 Fw-190 on July 2, 1943; Downed 1 Fw-190 on January 28, 1944; Downed 1 Me-109 on January 28, 1944"
```


```{r Create new variables}

#####Create long format with the victories --> ezt majd szerintem szét kellene szedni külön 2 chunkba
row.names(rdat2) <- 1:nrow(rdat2)
rlong <- as.data.frame(matrix(nrow=0,ncol=18))
#colnames(rlong) <- c("id","num","type","date")
cn <- c(c("id","num","type","date"),colnames(rdat2[,c(4:10,12:18)]))
colnames(rlong) <- cn
for(temprow in (1:nrow(rdat2))){
  temps <- rdat2[temprow,c(4:10,12:18)]
  tempid <- temprow
  tempvic <- rdat2[temprow,"aerial_victory_credits"]
  tempnum <- NA
  temptype <- NA
  tempdate <- NA
  if(!is.na(tempvic)){
    tempvic2 <- strsplit(tempvic, split="; ")
    for(vic in unlist(tempvic2)){
      vic2<-unlist(strsplit(unlist(vic),split=" "))
      if(length(vic2)==8){
        tempnum1 <- eval(parse(text=vic2[2]))
        tempnum2 <- eval(parse(text=vic2[3]))
        tempnum <- tempnum1 + tempnum2
        vic2 <- vic2[-3]
      }else{
        tempnum <- eval(parse(text=vic2[2]))
      }
      temptype <- vic2[3]
      tempdate <- as.Date(gsub(",","",paste(vic2[7], vic2[5], vic2[6],sep="-")), format = "%Y-%B-%d")
      temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
      colnames(temprow) <- cn
      rlong <- rbind(rlong,temprow)
    }
  }else{
    temprow <- cbind(data.frame(tempid, tempnum, temptype, tempdate),temps)
    colnames(temprow) <- cn
    rlong <- rbind(rlong,temprow)
  }
}
rlong$date <- as.Date(rlong$date)
```

```{r long format data exploration}
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  group_by(date) %>% 
  summarize(n=n())
rlong %>% 
  mutate(date = format(date, format = "%Y")) %>% 
  filter(hasvictory==T) %>% 
  ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1),y=as.numeric(format(date, format = "%Y"))+rnorm(nrow(rlong[rlong$hasvictory==T,]), 0, 0.1)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+scale_y_continuous(breaks = seq(1941,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
    ggplot(aes(x=as.numeric(format(gradyear, format = "%Y"))+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.1),y=number_of_aerial_victory_credits+rnorm(nrow(rdat2[rdat2$hasvictory==T,]), 0, 0.05)))+geom_point()+scale_x_continuous(breaks = seq(1942,1945,1))+theme_classic()
rdat2 %>% 
  filter(hasvictory==T) %>% 
  group_by(gradyear,rank_at_graduation,state) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n))
```

# Report

# Discussion