---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
#rd <- tidytuesdayR::tt_load('2020-09-22') #sadly this did not work, so I ended up downloading the datafile itself
exp<-read.table(url("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-09-22/expeditions.csv"), sep = ",", dec=".", header = T)


fakt <- as.data.frame(forcats::fct_lump(f=exp$peak_name,n=15))

ord<-exp %>% 
  mutate(peak_name = forcats::fct_lump(peak_name, n=15)) %>%
  filter(peak_name!="Other") %>% 
  group_by(peak_name) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  select(peak_name)
fct_reorder()

exp %>% 
  mutate(peak_name = forcats::fct_lump(peak_name, n=15)) %>%
  filter(peak_name!="Other") %>% 
  group_by(peak_name, season) %>% 
  summarize(ordi=n(), .groups="drop") %>% 
  mutate(peak_name = fct_reorder(peak_name,ordi,sum, .desc=FALSE)) %>%
  ggplot(aes(y = peak_name, x = ordi, fill = season))+geom_col()+scale_fill_viridis_d()+ylab("")+xlab("Number of expeditions")+theme_light()+theme(legend.position = "bottom")+ggtitle("The 15 most popular peaks stacked by season of expedition")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
#phdr<-read.table(url("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-02-19/phd_by_field.csv"), sep = ",", dec=".", header = T) first I used this
#but it had only 1685 lines, so I was forced to fix my issues with tidytuesdayR::tt_load
phdr <- tidytuesdayR::tt_load('2019-02-19') #fixed it, had to create a new token and add it to .Renviron as GITHUB_PAT
phdr <- as.data.frame(phdr[1])
colnames(phdr) <- c("broad_field","major_field","field", "year","n_phds")
phd1 <- phdr %>% 
  mutate(broad_field = as.factor(broad_field), major_field = as.factor(major_field))
phd1 %>% 
  group_by(broad_field,year) %>% 
  summarize(summi = sum(n_phds, na.rm=T)) %>%
  ggplot(aes(y=summi,x=year,color = broad_field))+geom_line(linewidth=1.2) + scale_x_continuous(breaks = scales::breaks_pretty(n=length(unique(phd1$year))))+theme_minimal()+scale_color_brewer(palette="Dark2", "Broad field") + ylab("") + xlab("") + ggtitle("Number of awarded Ph.D.-s in the US by year")
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commuter <- as.data.frame(tidytuesdayR::tt_load('2019-11-05')[1])
str(commuter)
summary(commuter)
commuted <- commuter %>% 
  mutate(commute.city = as.factor(commute.city), commute.state=as.factor(commute.state), commute.city_size = as.factor(commute.city_size), commute.mode = as.factor(commute.mode))
commuted %>% 
  group_by(commute.state, commute.state_region, commute.state_abb) %>% 
  pivot_wider(names_from = commute.mode, values_from = commute.n) %>% 
  summarize(sumbike = sum(Bike, na.rm = T), sumwalk = sum(Walk, na.rm=T)) %>% 
  ggplot(aes(x=sumwalk, y=sumbike, color=commute.state_region, label=commute.state_abb))+geom_point(size=2) + scale_y_log10(labels = scales::comma_format()) + scale_x_log10(labels = scales::comma_format()) + geom_text(color = "black") + theme_light() + xlab("Number of ppl walking to work (log N)") + xlab("Number of ppl biking to work (log N)")
  
asd <- split(x=commuted,f=commuted$commute.mode, drop = F)
asd2<-merge(as.data.frame(asd[1]), as.data.frame(asd[2]),by=2)
```
