---
title: "Assignment 3: Model comparison"
author: "Marton Kovacs"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.
```{r load in libraries, include=FALSE}
library(readxl)
library(dplyr)
library(tidyverse)
library(car)
library(lmtest)
library(lm.beta)
```

```{r load&basic exploration, include=FALSE}
filepath <- "/home/equarius/Documents/uni/semester_2/stat_programming/elte-ppk-r-course-Spring-2026/data/assignment_3_dataset.xlsx"
rd<-read_excel(filepath,sheet=1)
unique(rd$sex)
#crd <- rd %>% 
#  as.data.frame() %>% 
#  mutate(sex = as.factor(sex)) %>% 
#  mutate(sex = recode(sex, "woman" = "female")) #sex was corrected here, but car broke this
rd[rd$sex=="woman","sex"] <- "female"
crd <- rd %>% 
  mutate(sex = as.factor(sex))
str(crd)
summary(crd)
crdl <- crd %>% 
  select(!sex) %>% 
  pivot_longer(!ID, names_to="measure",values_to="value")
crdl$sex <- NA
for(tempid in unique(crdl$ID)){
  tempid <- as.character(tempid)
  tempsex <- as.numeric(crd[crd$ID==tempid,"sex"])
  crdl[crdl$ID==tempid,"sex"] <- tempsex
}
crdl$sex <- factor(as.character(crdl$sex), levels = c("1","2"), labels = c("1" = "female","2" = "male")) #delete this before submission
```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r EDA and recode, include=FALSE}
baseplot <- crdl %>% 
  ggplot(aes(y = value))+facet_wrap(~measure, scales = "free")
baseplot + geom_boxplot()
baseplot + geom_density()
crd %>% 
  summary()
crdl %>% 
  group_by(measure) %>% 
  summarize(mean = mean(value),
            sd = sd(value),
            min = min(value),
            max = max(value),
            median = median(value),
            shapiro = round(as.double(shapiro.test(value)[2]),2)) %>% 
  arrange(desc(mean))
```

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r corrections, include=FALSE}
#delete crdl here before submission
crd$mindfulness <- round(crd$mindfulness)
crd<-crd[!crd$pain>10,]
#see the correction of sex at line 72
crdl[crdl$measure=="mindfulness","value"]<-round(crdl[crdl$measure=="mindfulness","value"]) #correct for mindfulness, max value should be 6
crdl <- as.data.frame(crdl[!(crdl$measure=="pain"&crdl$value>10),]) #remove pain 50 outlier
```

**The following have been found**
+ sex included "woman" --> recoded to "female"
+ mindfulness had values above 6, which should have been the maximum.
+ pain had a value above the maximum of 10, this case was dropped

### Model diagnostics
#### Build the more complex model

*__more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__*


In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r __more complex model__, include=FALSE}
mc <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva+cortisol_serum, data = crd)
```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}
coodi <- cooks.distance(mc)
coodi %>% 
  as.data.frame() %>% 
  rownames_to_column(var="ID") %>% 
  mutate(ID = as.numeric(ID)) %>% 
  ggplot(aes(x=ID,y=coodi))+geom_bar(stat="identity")+geom_hline(yintercept=4/nrow(crd))+theme(axis.text.x = element_text(angle=90))+scale_x_continuous(breaks=seq(1,159,1))
#no element is D > 1, but a more rigorous 4/n correction shows some more influential observations
#creating a model with excluded cases
excl <- coodi %>% 
  as.data.frame() %>% 
  rownames_to_column(var="ID") %>% 
  filter(coodi > 4/nrow(crd)) %>% 
  select(ID)
excl <- as.numeric(unname(unlist(excl)))
crdwo <- crd[-excl,]
#re-creating model for later analysis
mcwo <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva, data = crdwo)
```

#### Checking assumptions

Check the normality assumption.

```{r}
residuals(mc) %>% 
  as.data.frame() %>% 
  ggplot(aes(sample=residuals(mc)))+stat_qq_line()+stat_qq() #QQ-plot doesn't indicate violation of normal distribution for residuals
```

Check the linearity assumption.

```{r}
plot(mc, which = 1)
#residual vs fitted plot doesn't show the violation of the assumption of linearity.
```

Check the homoscedasticty assumption (homogeneity of variance).

```{r}
plot(mc, which = 3)
bptest(mc)
#trendline and Breusch-Pagan test suggest homoscedasticity
```

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}
vif(mc)
cor(crd[,c(2, 4:12)], method="spearman") %>% #using spearman, because of violation of normal distribution at multiple vars, see at row 108
  as.data.frame() %>% 
  rownames_to_column(var = "ID") %>% 
  pivot_longer(!ID, names_to="measure",values_to="value") %>% 
  filter(value >=0.8) %>% 
  group_by(measure)
#The two cortisol measures are highly correlated
```


### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}
#I decided to drop the serum cortisol measure, to avoid violating the linearity assumption.
#I chose this, because multiple clinical results already suggest its connection to experienced pain levels, thus it would be optimal
#to explore the saliva level's associations further.
mcc <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva, data = crd)
```

#### Checking outliers of the updated model

```{r}
coodi <- cooks.distance(mcc)
coodi %>% 
  as.data.frame() %>% 
  rownames_to_column(var="ID") %>% 
  mutate(ID = as.numeric(ID)) %>% 
  ggplot(aes(x=ID,y=coodi))+geom_bar(stat="identity")+geom_hline(yintercept=4/nrow(crd))+theme(axis.text.x = element_text(angle=90))+scale_x_continuous(breaks=seq(1,159,1))
#Outliers are still present, but kept for natural variability
```

#### Checking assumptions of the updated model

Normality assumption

```{r}
residuals(mcc) %>% 
  as.data.frame() %>% 
  ggplot(aes(sample=residuals(mcc)))+stat_qq_line()+stat_qq()
#Normality assumptions still holds
```

Linearity assumption

```{r}
plot(mcc, which = 1)
#No visible change
```

Homoscedasticty assumption (homogeneity of variance)

```{r}
plot(mcc, which = 3)
bptest(mcc)
#Slightly larger edge-effect at the right tip, but the assumption still holds
```

Multicollinearity assumption

```{r}
vif(mcc)
#The issue of multicollinearity seems to be accounted for, none of the values are above 3
```

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

```{r}
ms <- lm(pain ~ age + sex, data = crd)
```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

*Report the results of the simpler model and the more complex model. For both models you should report the* **model test statistics (adj.R2, F, df, and p value)***. Also, report the statistics describing the* **coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).**

```{r}
mcc <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva, data = crd)
summary(mcc) #Complex model
#Residual standard error = 1.238
#adj.R2 = 0.34
#F(6,152)=14.53
#p<0.001
summary(ms) #Simple model
#Residual standard error = 1.464
#adj.R2 = 0.07
#F(6,156)=7.511
#p<0.001
msreport <- cbind(summary(lm.beta(ms))$"coefficients",confint(ms)) %>% 
  as.data.frame() %>% 
  rownames_to_column("predictor") %>% 
  select(!4&!5)
colnames(msreport) <- c("predictor", "B", "Beta","p","confind_lower","confint_upper")
mccreport <- cbind(summary(lm.beta(mcc))$"coefficients",confint(mcc)) %>% 
  as.data.frame() %>% 
  rownames_to_column("predictor") %>% 
  select(!4&!5)
colnames(mccreport) <- c("predictor", "B", "Beta","p","confind_lower","confint_upper")
```

|model|Residual_standard_error|adj.R2|F|df|p
| -------- | -------- | -------- | -------- | -------- | -------- |
|simple|1.46|0.07|7.51|6,156|<0.001|
|complex|1.24|0.34|14.53|6,152|<0.001|

Our model equation will look like:

Ypain = 1.47(intercept) - 0.23(age) + 0.31(male sex) - 0.03(STAI) + 0.09(pain catastrophization) - 0.14(mindfulness) + 0.58(saliva cortisol)

*Compare the two models in terms of how much variance they explain of pain’s variability in the sample.***Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test** *comparing the two models.*

Compare the two models.

```{r echo=FALSE}
anova(ms, mcc)
#F(4,152)=16.536, p<0.001
AIC(ms) #577.50
AIC(mcc) #528.05
lrtest(ms,mcc) #p<0.001
summary(lm.beta(ms))
summary(lm.beta(mcc))
```

### Comparing the two models

The simple baseline model containing only age and sex as predictors explained a significant, but small portion of variability R2=0.09, adjusted R2=0.07, F(2,156)=7.51, p<0.001, AIC = 577.50.
The more complex model containing trait anxiety, pain catastrophization, mindfulness and saliva cortisol explained a significant, but larger portion of variability R2=0.36, adjusted R2=0.34, F(6,152)=14.53, p<0.001, AIC = 528.05.
R2 values indicate more variance explained with the full model and the lower AIC value also point to a better model fit. Hierarchical F-test indicate that the complex model improved the model explanation (F(4,152)=16.536, p<0.001) and the significance of the likelihood ratio test also point into this direction (p<0.001).

# Discussion

Examining the predictor performances, in the simple model age was the only significant predictor (p<0.001), where every year would mean a 0.09 points of decrease in the pain. This would assume that - in line with previous findings - pain has a protective effect.
If we examine the complex model, the pain catastrophization and the cortisol saliva remained only significant (both p<0.001). This would assume that behind the protective effects of age in reality we see psychological traits. With every increase in pain catastrophization the pain would increase 0.09 points; and with every increase in cortisol saliva the pain would increase with 0.58 points, pointing to the fact that cortisol saliva taken 5 minutes before the operation was the best predictor of the experienced level of pain after the operation.

The addition of psychological and physiological variables improved our understanding about the nature of postoperative pain. It also helped in untangling the association with age.